---
title: "MDSR Capstone - Interactive Earthquake Plots"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(MSDRCapstone)
```

Package: MSDRCapstone - completed as part of Coursera Mastering Software Development
with R Specialization.

GitHub:  https://github.com/SondergardM/MSDRCapstone

## Data Source

The functions in this R package utilize data from the US National Oceanic and 
Atmospheric Administration's (NOAA's) National Center for Environmental Information
(NCEI) [Global Significant Earthquake Database](https://www.ngdc.noaa.gov/hazard/earthqk.shtml),
which is a covers significant earthquakes from 2150 B.C. to the present.  For the purposes
of this package, data was downloaded from the year 2000 A.D. to the present and has been
included in the package (data set **noaa_data**).  The data set includes 1219 observations and 39 fields.

## Package Summary

The functions in this package are designed to utilize data from NOAA's earthquake 
database and produce two basic sets of plots: (1) the first is an earthquake timeline
for a given country (or countries) that shows the date and magnitude of the earthquake,
along with data labels for the five most significant quakes in the data chosen by
the user; (2) an HTML map showing the geographic locale of the earthquake along with
either a date or other related data shown as a pop-up label when the user clicks on a
given earthquake location.

The package consists of six (6) functions:

* **eq_location_clean()** - Function strips out country name for data downloaded from 
 the NOAA earthquake database and converts the column names to title case for readability.

    Example: 
    
```{r eval = FALSE}
      earthquakes <- eq_location_clean(noaa_data)
```      

* **eq_clean_data()** - Function combines Year, Month and Day fields into a single date variable
 and prepares the data for use in plots.

    Example: 
 
```{r eval = FALSE} 
      earthquakes <- eq_clean_data(earthquakes)
```

* **geom_timeline()** - Function plots a time line for earthquakes ranging from "xmin" to "xmax"
 dates, with a point for each earthquake.  The x aesthetic is a date and the y aesthetic is a
 factor indicating some stratification, in which case multiple time lines will be plotted for
 each level of the factor (e.g. country).
 
    Example:
```{r eval = FALSE}
 
      earthquakes %>% dplyr::filter(Country %in% c("Afghanistan")) %>%  
                dplyr::filter(Date_ymd > "2000-01-01") %>%  
                ggplot() +  
                      geom_timeline(aes(x = Date_ymd,  
                                        y = Country,  
                                        colour = Total_Deaths,  
                                        size = EQ_Primary)) +  
                      scale_size_continuous(name = "Richter Scale Magnitude") +  
                      scale_colour_continuous(name = "Total Deaths",  
                                              type = "viridis") +  
                      theme_light() +  
                      xlab("Year")  
```
  
* **geom_timeline_label()** - Function adds annotations to the earthquake data provided by the
 geom_timeline() function.  The geom adds a vertical line to each data point with a text annotation
 providing the location of the earthquake. The x aesthetic is a date of the earthquake and a label
 which takes the column name from which annotations will be obtained.
 
  Example:
  
```{r eval = FALSE}
      earthquakes %>%  
        dplyr::filter(Country %in% c("China", "New Zealand")) %>%  
        dplyr::filter(Date_ymd > "2000-01-01") %>%  
        dplyr::select(Date_ymd, Country, Location_Name, Latitude,  
                 Longitude, EQ_Primary, Total_Deaths) %>%  
        drop_na(Location_Name) %>%  
        ggplot() +  
          geom_timeline(aes(x = Date_ymd,  
                            y = Country,  
                            colour = Total_Deaths,  
                            size = EQ_Primary)) +  
          geom_timeline_label(aes(x = Date_ymd,  
                                  magnitude = EQ_Primary,  
                                  label = Location_Name,  
                                  y = Country,  
                                  n_max = 5)) +  
        scale_size_continuous(name = "Richter Scale Magnitude") +  
        scale_colour_continuous(name = "Total Deaths", type = "viridis") +  
        theme_light() +  
        xlab("Year")
```
  
* **eq_map()** - Function creates an interactive map showing the location of earthquakes in the given
 earthquakes data set.  The size of the circles in the map are proportional to the magnitude of
 the earthquakes. The map is interactive, and when you click on a link, a popup label shows the
 annotation as specified by the **annot_col** variable.
 
  Example:
 
```{r eval = FALSE} 
      earthquakes <- earthquakes %>%  
                      dplyr::filter(Country == 'China') %>%  
                      dplyr::filter(lubridate::year(Date_ymd) >= 2000)  
      eq_map(earthquakes, annot_col = 'Date_ymd')  
```
  
* **eq_create_label()** - Function creates a more descriptive and HTML-formatted popup
 label to be used with the **eq_map()** function. This function creates a vector of HTML-formatted
 labels using supplied data from the eathquakes data set.  The label contains the following
 information:

    * Date in YMD format
    * Location Name (as cleaned by the **eq_location_clean()** function)
    * Earthquake magnitude (Richter scale)
    * Total Deaths caused by the event

  Example:
  
```{r eval = FALSE}
      earthquakes <- earthquakes %>%  
                   dplyr::filter(Country == 'China') %>%  
                   dplyr::filter(lubridate::year(Date_ymd) >= 2000) %>%  
                   dplyr::mutate(Popup_Text = eq_create_label(.))  
      eq_map(earthquakes, annot_col = "Popup_Text")  
```
   
Further documentation can be found in the package documentation for each individual
function.
